---
title: "Project 1"
author: "Yandong Xiong"
date: "1/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages needed
require(tidyverse)
require(haven)
require(kableExtra)
```

# Introduction

This project is to look deeper into the patterns and characteristics of voter behaviors to the 2016 Election. The dataset, downloaded from ANES website, was designed and collected by the ANES Principal Investigators and staff with design input from the ANES advisory board and contributors to the ANES Online Commons. 

The initial part is "Data Exploring", which includes import, manipulations and basic view on the raw dataset. Then, "Inspection" for each variable in the studied dataset. Finally, "Visualization" on properties and patterns of the voters.


# Data exploring

```{r Manipulation}
# change the working directory to the data file location
setwd("~/Documents/GitHub/Spring2021-Project1-yx2659/data") 

# read the dataset
dt <- read_sav("anes_timeseries_2016.sav")

# basic lookup for the raw dataset
head(dt)
dim(dt)

# manipulations for research questions
# rename the variables
data <- dt %>% rename(ID = V160001, gender = V161002, interview_mode = V160501, state = V161010e, registration_turnout_status = V162065x, 
                      education = V161270, race = V161310x, presidential_vote = V162062x, House_vote = V162067x, Senate_vote = V162068x,
                      Governer_vote = V162069x
                      ) %>%
  # create and modify new variables
  mutate(registration_turnout_status = ifelse(registration_turnout_status == 0, "not registered, not voted",
                                              ifelse(registration_turnout_status == 1, "registered, not voted",
                                                     ifelse(registration_turnout_status == 2, "registered, voted", NA))),
         presidential_vote = ifelse(presidential_vote == 1, "Clinton",
                                    ifelse(presidential_vote == 2, "Trump",
                                           ifelse(presidential_vote == 3, "Johnson",
                                                  ifelse(presidential_vote == 4, "Stein",
                                                         ifelse(presidential_vote == 5, "others", "no vote"))))),
         House_vote = ifelse(House_vote == 1, "Democratic",
                             ifelse(House_vote == 2, "Republican",
                                    ifelse(House_vote == 3, "others", "no vote"))),
         Senate_vote = ifelse(Senate_vote == 1, "Democratic",
                             ifelse(Senate_vote == 2, "Republican",
                                    ifelse(Senate_vote == 3, "others", "no vote"))),
         Governer_vote = ifelse(Governer_vote == 1, "Democratic",
                             ifelse(Governer_vote == 2, "Republican",
                                    ifelse(Governer_vote == 3, "others", "no vote"))),
         gender = ifelse(gender == 1, "male", ifelse(gender == 2, "female", NA)), 
         interview_mode = ifelse(interview_mode == 1, "FTF", "Web"),
         race = ifelse(race == 1, "white", 
                       ifelse(race == 2, "black",
                              ifelse(race == 3, "Asian",
                                     ifelse(race == 4, "Native American or Alaska Native",
                                            ifelse(race == 5, "Hispanic", 
                                                   ifelse(race == 6, "other", NA)))))),
         education  = ifelse(education %in% 1:8, "less than high school graduate", 
                             ifelse(education %in% c(9,90), "high school graduate", 
                                    ifelse(education == 10, "college but no degree",
                                           ifelse(education %in% c(11,12), "associate college degree", 
                                                  ifelse(education == 13, "bachelor's degree",
                                                         ifelse(education == 14, "master's degree",
                                                                ifelse(education == 15, "professional school degree", "doctoral degree")))))))
         ) %>%
  # drop rows that respondents do not want to share the information for vote registration status and races or respond not avaliable
  filter(!is.na(registration_turnout_status) & !is.na(race)) %>%
  # select nedded variables useful for the research questions
  select(ID, interview_mode, gender, state, education, race, registration_turnout_status, presidential_vote, House_vote, Senate_vote, Governer_vote)

# basic lookup on the manipulated dataset
head(data, n=10)
dim(data)
```

After manipulations on the raw data, I selected 10 variables to be studied, shown in the following table: 

```{r}
text_tbl <- data.frame(
Variable = c("ID", "interview_mode", "gender", "state", "education", "race", 
              "registration_turnout_status", "presidential_vote", "House_vote",  "Senate_vote", "Governer_vote"), 
Type = c("Quantitative", rep("Qualitative", 10)),
Description = c("Case ID", "Mode of interview", "gender of interviewers for FTF mode only", "2-char state Postal abbreviation for sample address", "Highest level of Education", "self-identified race", "2016 registration - turnout status", "2016 Presidential vote", "party of 2016 U.S. House vote", "party of 2016 U.S. Senate vote", "party of 2016 Governor vote"))
kbl(text_tbl, booktabs = T) %>% kable_styling(full_width = F) %>% column_spec(1, bold = T, color = "red")
```

# Inspection

Some inspections on each variable based on their distributions:

```{r Inspections}
# Inspections for each variables 
(mod = data %>% count(interview_mode) %>% mutate(propotion = round(n/sum(n),2)))
pie_mod = ggplot(mod, aes(x = "", y = propotion, fill = interview_mode)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_mod

(gen = data %>% count(gender) %>% mutate(propotion = round(n/sum(n),2)))
pie_gen = ggplot(gen, aes(x = "", y = propotion, fill = gender, na.rm = FALSE)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_gen

(sta = data %>% count(state, sort = TRUE) %>% mutate(propotion = round(n/sum(n),2)))
pie_sta = ggplot(sta, aes(x = "", y = propotion, fill = state)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_sta

(rac = data %>% count(race, sort = TRUE) %>% mutate(propotion = round(n/sum(n),2)))
pie_race = ggplot(rac, aes(x = "", y = propotion, fill = race)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_race

(edu = data %>% count(education, sort = TRUE) %>% mutate(propotion = round(n/sum(n),2)))
pie_edu = ggplot(edu, aes(x = "", y = propotion, fill = education)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_edu

(vote_status = data %>% count(registration_turnout_status, sort = TRUE) %>% mutate(propotion = round(n/sum(n),2)))
pie_vote_status = ggplot(vote_status, aes(x = "", y = propotion, fill = registration_turnout_status)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_vote_status

(pre_vote = data %>% count(presidential_vote, sort = TRUE) %>% mutate(propotion = round(n/sum(n),2)))
pie_pre_vote = ggplot(pre_vote, aes(x = "", y = propotion, fill = presidential_vote)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_pre_vote

(hou_vote = data %>% count(House_vote, sort = TRUE) %>% mutate(propotion = round(n/sum(n),2)))
pie_hou_vote = ggplot(hou_vote, aes(x = "", y = propotion, fill = House_vote)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_hou_vote

(gov_vote = data %>% count(Governer_vote, sort = TRUE) %>% mutate(propotion = round(n/sum(n),2)))
pie_gov_vote = ggplot(gov_vote, aes(x = "", y = propotion, fill = Governer_vote)) +
  geom_bar(stat = "identity") + 
  coord_polar("y") + theme_void()
pie_gov_vote
```




```{r Visualization}
df1 = data %>% group_by(presidential_vote) %>% mutate(count = n())
ggplot(df1) +
  geom_bar(mapping = aes(x = reorder(presidential_vote, -count), fill = registration_turnout_status)) +
  coord_flip() + 
  labs(
    title = "",
    x = ""
  )

df2 = data %>% group_by(education) %>% mutate(count = n())
ggplot(df2) +
  geom_bar(mapping = aes(x = reorder(education, -count), fill = registration_turnout_status)) +
  coord_flip() + 
  labs(
    title = "",
    x = ""
  )

df3 = data %>% group_by(race) %>% mutate(count = n())
ggplot(df3) +
  geom_bar(mapping = aes(x = reorder(race, -count), fill = registration_turnout_status)) +
  coord_flip() + 
  labs(
    title = "",
    x = ""
  )
```

